---
title: "04-Shared_Computing_Resource_Etiquette"
output: html_document
---
```{r, include = FALSE}
leanbuild::set_knitr_image_path()
```

# Shared Computing Etiquette

In this chapter we will discuss the proper etiquette for using shared computing resources. This will help you to understand what would be required for you to use such resources. Different resources will have slightly different use rules, however, many will share common usage requirements.


```{r, fig.align='center', echo = FALSE, fig.alt= "Learning Objectives: 1.Describe the size of various data, 2. Recognize the difference between personal computers and remote computing options. 2.Explain how cloud computing and server computing works 3. Determine which options will provide enough computing capacity for your research needs", out.width="100%"}
leanbuild::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf5f8818810_1_171")
```

We will use the Johns Hopkins Joint High Performance Computing Exchange (JHPCE) is a High-Performance Computing (HPC) cluster facility as an example for guidance.

For this particular resource there are 400 active users.It is optimized for genomic and biomedical research and has 4,000 cores! That's right, as you can imagine, this is much more powerful than the individual laptops and desktops that researchers at the university have for personal use, which would typically currently only have around 8 cores. There is also 28TB of RAM and 14 PB of storage!! Now that you know more about digital sizes, you can appreciate that this server can allow for much faster processing and really large amounts of storage, as again a researchers' computer might have something like 16 GB of RAM and 1TB of storage. Individual users connect and perform jobs on the cluster using a formal common pool resource (CPR) hierarchy system. 

## General rules

 - don't share password
 - don't access a server on a computer that is not authorized to do so
 - don't install software unless you have permission
 - don't use the server for storage or computation that you are not authorized for
(personal use)
 - don't alter configuration or security settings without authorization
 - immediately report any data security concerns
 - Don't use more nodes if you don't need to (if you need all nodes - let someone know!)
 - Don't use all RAM on a node if you don't need to (if you need to - let someone know!)
 - Communicate with others if you will be submitting a large or intensive job
 - Don't submit too many jobs at once

https://www.doi.gov/sites/doi.gov/files/uploads/DOI-RULES-OF-BEHAVIOR-FOR-COMPUTER-NETWORK-USERS-REFERENCE-GUIDE.pdf

## Running Jobs

### Specifying memory (RAM) needs

### Checking status




