

```{r, include = FALSE}
ottr::set_knitr_image_path()
```

# Research Platforms

In this chapter we will provide examples of computing platforms that are designed to help researchers and that you might find useful for your work. Please note that we aim to provide a general overview of options and thus this is not a complete list. Let us know if there is a platform or system that you think we should include!

```{r, fig.align='center', echo = FALSE, fig.alt= "Learning Objectives: 1.Recognize general informatics computing and data management platforms. 2.Be aware of differnt benefits of the various platforms", out.width="100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf5f8818810_1_171")
```


The major advantage of these platforms is that users can analyze data where it lives, as many platforms host public data. However, some also allow you to upload your own data.  There is less need for data transfers, as you can analyze your data, store your data and share it in one place, saving time. Users can sometimes also share how they did their analysis as well, improving reproducibility practices. Additionally, another advantage is that some of these platforms also provide educational material on how to work with data. 

### Galaxy

This section was written by [Jeremy Goecks](https://goeckslab.org/people/jeremy.html):

Galaxy is a web-based computational workbench that connects analysis tools, biomedical datasets, computing resources, a graphical user interface, and a programmatic API. Galaxy (https://galaxyproject.org/) enables accessible, reproducible, and collaborative biomedical data science by anyone regardless of their informatics expertise. There are more than 8,000 analysis tools and 200 visualizations integrated into Galaxy that can be used to process a wide variety of biomedical datasets. This includes tools for analyzing genomic, transcriptomic (RNA-seq), proteomic, metabolomic, microbiome, and imaging datasets, tool suites for single-cell omics and machine learning, and thousands of more tools. Galaxy’s graphical user interface can be used with only a web browser, and there is a programmatic API for performing scripted and automated analyses with Galaxy.

```{r, fig.align='center', echo = FALSE, fig.alt= "Galaxy can be accessed through a web browser and provides users with access to tools, datasets, computing resources, a graphical user interface (GUI) for users who would like to interact with Galaxy by clicking buttons and using drop-down menus and a programmtic API for users that would like to write code to interact with Galaxy", out.width= "100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_131")
```

Galaxy is used daily by thousands of scientists across the world. A vibrant Galaxy community has deployed hundreds of Galaxy servers across the world, including more than 150 public and three large national/international servers in the United States, Europe, and Australia (https://usegalaxy.org, https://usegalaxy.eu, https://usegalaxy.org.au). The three national/international servers have more than 250,000 registered users who execute >500,000 analysis jobs each month. Galaxy has been cited more than 10,000 times with >20% from papers related to cancer. The Galaxy Tool Shed (https://usegalaxy.org/toolshed) provides a central location where developers can upload tools and visualizations and users can search and install tools and visualizations into any Galaxy server. Galaxy has a large presence in the cancer research community. Galaxy serves as an integration and/or analysis platform for 7 projects in the NCI ITCR program. There is also increasing use of Galaxy in key NIH initiatives such as the NCI Cancer Moonshot Human Tumor Atlas Network (HTAN) and the NHGRI Data Commons, called the AnVIL (https://anvilproject.org/).

```{r, fig.align='center', echo = FALSE, fig.alt= "Map of the 3 Galaxy servers", out.width= "100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_135")
```


Galaxy’s user interface, accessible via a web browser, provides access to all Galaxy functionality. The main Galaxy interface has three panels: available tools (left), running analyses and viewing data (middle), and a full history of tools run and datasets generated (right). 

```{r, fig.align='center', echo = FALSE, fig.alt= "Galaxy user interface showing the 3 panels: left = tools, middle = analysis and data, and right = history and generated datasets", out.width= "100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_111")
```


Datasets for analysis in Galaxy can be **uploaded** from a laptop or desktop computer or obtained from public data repositories connected to Galaxy. With Galaxy, complex workflows composed of tens or even hundreds of analysis tools can be created and run. In Galaxy’s workflow interface, tools can be added and connected via a simple drag-and-drop approach. 

```{r, fig.align='center', echo = FALSE, fig.alt= "Galaxy workflow example showing the beginning steps of a workflow", out.width= "100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfb2e21ecdc_0_141")
```


Galaxy users can share all their work—analysis histories, workflows, and visualizations—via simple URLs that are available to specific colleagues or a link that anyone can access. Galaxy’s user interface is highly scalable. Tens, hundreds, or even thousands of datasets can be grouped into collections and run in parallel using individual tools or multi-tool workflows. In summary, Galaxy is a popular computational workbench with tools and features for a wide variety of data analyses, and it has broad usage in cancer data analysis.

See [here](https://toolshed.g2.bx.psu.edu/) for the list of applications supported by Galaxy and [here](https://training.galaxyproject.org/) for more information on how to use Galaxy resources.


### Terra

[Terra](https://terra.bio/) is a biomedical research computing platform that is based on the Google Cloud platform, that also allows users easier ways to manage the billing of their projects. It provides users with access to data, workflows, interactive analyses using Jupyter Notebooks, RStudio, and Galaxy, data access tools from [FireCloud from the Broad Institute](https://firecloud.terra.bio/) (a [National Cancer Institute (NCI)](https://www.cancer.gov/) Cloud Resource project),  as well as workspaces to organize projects and collaborate with others. Terra also has [many measures](https://terra.bio/resources/security/) to ensure that data is especially secure and the offer clinical features for ensuring that [health data is protected](https://terra.bio/about/privacy/). Note that according users who do upload protected health information must select to use these extra clinical features and enter a formal agree with [FireCloud](https://firecloud.terra.bio/) about their data. 

Importantly users can get access to use [Genotype -Tissue Expression (GTEx)](https://gtexportal.org/home/), [Therapeutically Applicable Research to Generate Effective Treatments (TARGET)](https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs000218.v24.p8) and [The Cancer Genome Atlas (TCGA)](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga) data using the platform. See [here](https://support.terra.bio/hc/en-us/articles/4402326091675-Accessing-GTEx-TARGET-TCGA-data) for information on how. 

Users can pay for data storage and computing costs for Google Cloud conveniently through Terra. Users can browse data for free.

Check out this video for more information:

```{r, fig.align="center", fig.alt = "video", echo=FALSE, out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/3rH86vcAqK8")
```



## AnVIL

If you could use some guidance on how to perform analyses using Galaxy and Terra, especially for genomic research, check out [AnVIL](https://anvilproject.org/), the [National Human Genome Research Institute (NHGRI)](https://www.genome.gov/) Analysis Visualization and Informatics Lab-space. It also provides access to many important genomic and related [datasets](https://anvilproject.org/data) from the NHGRI.


According to their website:

> By providing a unified environment for data management and compute, AnVIL eliminates the need for data movement, allows for active threat detection and monitoring, and provides elastic, shared computing resources that can be acquired by researchers as needed.

It relies on Terra for the cloud based compute environment, Dockstore for  standardized tools and workflows, Gen3 for data management for querying and organizing data, Galaxy tools and environment for analyses with less code requirements, and [Bioconductor](https://www.bioconductor.org/) tools for R programming users. [Bioconductor](https://www.bioconductor.org/) is a project with the mission to catalog, support, and disseminate bioinformatics open-source R packages. Packages have to go through a review process before being included. 


## CyVerse

[CyVerse](https://cyverse.rocks/about) is a computing platform  similar to Galaxy that also offers computing resources for storing, sharing, and working with data with a graphical interface, as well as an API. Computing was previously offered using the cloud computing platform from CyVerse called [Atmosphere](https://cyverse.org/refocusing-atmosphere-to-support-cloud-native-development), which relied on users using virtual machines. Users will now use a new version of Atmosphere with partnership with [Jetstream](https://jetstream-cloud.org/). This allows users to use containers for easier collaboration and also offers US users more computing power and storage. Originally called iPlant Collaborative, it was started by a funding from the National Science Foundation (NSF) to support life sciences research, particularly to support ecology, biodiversity, sustainability, and agriculture research. It is led by the University of Arizona, the Texas Advanced Computing Center, and Cold Spring Harbor Laboratory. It offers access to an environment for performing analyses with Jupyter (for Python mostly) and RStudio (for R mostly) and a variety of tools for Genomic data analysis. See [here](https://cyverse.atlassian.net/wiki/spaces/DEapps/pages/241882146/List+of+Applications) for a list of applications that are supported by CyVerse.  Note that you can also install tools on both platforms. Both CyVerse and Galaxy offer lots of helpful documentation, to help users get started with informatics analyses.

See [here](https://learning.cyverse.org/en/latest/) to learn more.

```{r, fig.align='center', echo = FALSE, fig.alt= "CyVerse graphical interface for performing analyses", out.width= "100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd56752f25_0_0
")
```

## SciServer

SciServer is also similar to Galaxy in that it is accessible through a web browser and allows users to store, upload, download, share, and work with data and common tools on the same platform. It was originally built for the astrophysics community (and called SkyServer) but it has now been adapted to be used by scientists of all fields and is indeed used by many in the genomics field.  It allows users to use Python and R in environments like Jupyter notebooks and RStudio, and also supports (Structured Query Language) SQL for data querying and management and is built on the use of Docker. 

The main idea of SciServer, is based on this premise: "bring the analysis to the data". It is free to use after users register. However, users can buy extra resources. Users can keep data private or share their data. 

As compared to Galaxy, this resources may be better for users with a bit more familiarity with informatics but who require more flexibility, particularly for working with collaborators such as physicists or material scientists as there are more tools supported across disciplines. In addition it also gives users access to very large data sets on Petabyte-scale (note that some of these require special permission to use) and supports developers to create their own web interfaces called SciUIs for particular use cases.

For @sciserver_2020 for more information.  


```{r, fig.align='center', echo = FALSE, fig.alt= "The SciServer dashboard allows users to interact with the various resources and tools available through SciServer. Most are available through the command line but some have user interface options. Jupyter Notebooks, RStudio and SciUIs (interfaces created by other SciServer users) can be used to interactively explore and analyze data by users.", out.width= "100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd56752f25_0_13")
```



## Materials Cloud

Another resource that might be of interest to Python users, particular those who collaborate with material scientists, is Materials Cloud. It is designed to promote reproducible work, collaboration, and sharing of resources among scientists, particularly for simulations for the materials science field. Users can share data in a citable way, download data, upload data, share workflows, and perform analyzes.

```{r, fig.align='center', echo = FALSE, fig.alt= "Materials Cloud resources are based on allowing users to Learn about resources, Work using the resources, Discover aspects about data that is available, Explore data with interactive graphs, and archive to store and share data.", out.width= "100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd56752f25_0_6")
```

This resource uses [AiiDAlab](https://www.materialscloud.org/work/aiidalab) as the computing environment for researchers, which is based on [AiiDA](https://www.nature.com/articles/s41597-020-00638-4). According to their website:

> AiiDAlab builds on [AiiDA](https://www.nature.com/articles/s41597-020-00638-4) as the computational workflow engine, and the Jupyter environment (notebooks, widgets, …) for writing and sharing apps

See [here](https://www.sciencedirect.com/science/article/pii/S092702562030656X?via%3Dihub) to learn more about how AiiDAlab supports the sharing of scientific workflows, particularly for those that use Python.

To learn more about Materials Cloud, check out @talirz_materials_2020.


## Overture

Overture is a relatively new option for perform large-scale genomic data analyses. You can upload, download, manage, analyze and share your data with authentication and authorization methods to add security. Although designed for genomic research, the [data management system](https://www.overture.bio/documentation/dms/) can be used for other scientific domains. Currently, additional products are still being developed for analysis, visualization, and sharing. However, several collaborations have created new incredible resources using some of the existing and developing products that might be useful for your research. Alternatively, Overture has options to help you create your own platform, see [here](https://www.overture.bio/services/) for more information. It is compatible with Google, Microsoft Azure and PostgreSQL for storage options. 

These collaborations using Overture products can be found on the [case studies](https://www.overture.bio/case-studies/) page of the [Overture website](https://www.overture.bio/).

For example, the [Cancer Genome Collaboratory](https://cancercollaboratory.org/) is one such collaboration. This is A cloud-based resource that allows researchers to perform analyses using [International Cancer Genome Consortium (ICGC)](https://en.wikipedia.org/wiki/International_Cancer_Genome_Consortium) cancer genome data, which includes tumor mutation data from the [The Cancer Genome Atlas (TCGA)](https://en.wikipedia.org/wiki/The_Cancer_Genome_Atlas) and the [Pan-Cancer Analysis of Whole Genomes (PCAWG)](https://dcc.icgc.org/pcawg) mutation data. See [here](https://cancercollaboratory.org/services-cloud-resources) for information about billing, storage capacity, access, and security. 

In addition, Overture products have also been used to create other data resources, such as the [Kids First Data Resource Portal](https://portal.kidsfirstdrc.org/login)  which has childhood cancer and birth defect genomic data for over 76,000 samples, and the [National Cancer Institute's Genomic Data Commons Data portal](https://portal.gdc.cancer.gov/), which also includes [The Cancer Genome Atlas (TCGA)](https://en.wikipedia.org/wiki/The_Cancer_Genome_Atlas) and [Therapeutically Applicable Research to Generate Effective Treatments (TARGET)](https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs000218.v24.p8). The portal supports some basic [analyses]((https://portal.gdc.cancer.gov/analysis) as well for clinical data statistics and survival analysis. 

## Globus

Globus provides developers of new platforms to manage, transfer, and share data with special attention to privacy and security. 

It has been used for several platforms such as the  [Systems Biology Knowledgebase (KBase)](https://www.kbase.us/), which is focused on integrating data across plants and microbes, [Biomedical Research Informatics Network (BIRN)](https://en.wikipedia.org/wiki/Biomedical_Informatics_Research_Network), which is a collaborative project to bring biomedical researchers together and share resources and data [@helmer_enabling_2011], and [Globus Genomics](https://www.globusgenomics.org/genomics/about.html),which uses the Globus data management infrastructure, Amazon web services, and Galaxy workflows to assist researchers with their genomics research endeavors. 


## Additional more specific platforms

### BaseSpace Sequence Hub

[BaseSpace](https://basespace.illumina.com/) is a platform that allows for data analysis of Illumina sequencing data and syncs easily with any Illumina sequencing machines that you might work with. There are many [applications](https://www.illumina.com/products/by-type/informatics-products/basespace-sequence-hub/apps.html) available to help you with your genomics research. They offer a 30 day free trial.

### GenePattern

[GenePattern](https://www.genepattern.org/) is similar to Galaxy in that it provides a web-based interface for genomic analyses. You can upload your own data, use workflows and pipelines form others and more! 

See [here](https://www.genepattern.org/user-guide) to access their user guide and [here](https://notebook.genepattern.org/quickstart/) for a quick start guide to using [GenePattern Notebook](https://notebook.genepattern.org/) which uses [Jupyter Notebooks](https://jupyter.org/) and GenePattern analysis tools to easily create data analysis reports. Users can also publish and share their notebooks with collaborators or the field, as well as access other people's notebooks that they can adapt for their own uses. See [here](https://notebook.genepattern.org/library/) for a collection of available notebooks. 


### ATLAS.ti

[ATLAS.ti](https://atlasti.com/) is designed particularly for qualitative analysis. You can use a variety of data types including video, audio, images, surveys, and social media data. A variety of tools, particularly for text data analysis are provided for methods such as [sentiment analysis](https://en.wikipedia.org/wiki/Sentiment_analysis), which is the process of assigning a general tone or feeling to text and [named-entity recognition](https://en.wikipedia.org/wiki/Named-entity_recognition), which is the process of extracting certain characteristics from texts that are what is called a [named entity] or a real-world object - such as a person's name or address. Such analyses can be helpful for understanding behaviors that might be associated with cancer risk. Although this type of analysis can be performed using R or Python among other coding languages, ATLAS.ti offers a nice graphical user interface to perform these types of analyses.Furthermore ATLAS.ti offers a great deal of flexibility about such analyses using different data types easily.

```{r, echo = FALSE, fig.alt= "Cheat sheet of Unix commands at https://files.fosswire.com/2007/08/fwunixref.pdf"}
knitr::include_url(url = "https://downloads.atlasti.com/docs/branding/atlasti_brochure_v9_EN_interactive_202110.pdf")
```



### XNAT

[XNAT](https://www.xnat.org/about/) offers computing resources and tools for performing imaging analysis and for storing and sharing imaging data in a HIPAA complaint manner (more on that in the coming). Developed by the [Bukner lab](https://cnl.rc.fas.harvard.edu/) previously at the Washington University and now at Harvard, it supports a variety of imaging data as well as other data types like clinical data.  Some tools can be used with a graphical interface and others with the command-line. See [here](https://wiki.xnat.org/documentation/case-studies) for example use cases. There is also a great deal of documentation available about how to use the tools and resources available at https://wiki.xnat.org/documentation.

```{r, fig.align="center", fig.alt = "video", echo=FALSE, out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/ENk589mOkhI")
```


### OHIF

The [open health imaging foundation (OHIF)](https://ohif.org) is a web-based imaging analysis platform that is widely used, particularly for radiology analysis, but it also supports whole-slide microscopy image analysis. It was developed by [Gordon Harris](https://ohif.org/team/) et al. and can be used for a variety of applications from cardiology to veterinary medicine. Check out these [example use cases](https://ohif.org/examples) of OHIF. OHIF also provides thorough documentation with images and videos about how to use the image viewer and tools available.


```{r, fig.align='center', echo = FALSE, fig.alt= "The OHIF image viewer (center) includes a tool bar for to modify the viewing and annotation of images (above center), a menu on the left to access images related to the current patient in other studies, and a menu on the right for making measurements and exporting measurement reports.", out.width= "100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd8d3f477a_159_2
")
```

For those interested, Gordon Harris and others are also working on a project called [Cornerstone](https://docs.cornerstonejs.org/), with the goal of providing software for others to display medical images in web browsers.


### PRISM

The Platform for Imaging in Precision Medicine called PRISM allows users to work with the vast data available from r the Cancer Imaging Archive (TCIA).

According to Fred Prior:

>It is designed to collect, curate and manage Radiology and Pathology images, clinical information associated with those images, annotations and image derived feature sets.  PRISM is designed to run on a Kubernettes cluster, but all of the components are containerized so they can run stand-alone or in an alternate orchestration framework.


See this [article](https://ascopubs.org/doi/full/10.1200/CCI.20.00001) for more information.
 



## Conclusion

We hope that this chapter has given you some more perspective on how the various computing options available designed for researchers like you. We also hope that you may have learned about another platform that can help you to make your research faster and more flexible.

In conclusion, here are some of the major take-home messages:

1) Computing platforms are cyberinfastructures that allow you to perform analyses with existing data and or allow you to upload, work with, and share your own data. 

2) There are quite a few platforms that are well developed with in depth documentation to support research quite broadly and allow users to work with a diverse set of data.

3) There are also platforms that were designed for specific research needs. 




