---
title: 'Specific Platforms'
output: html_document
---

```{r, include = FALSE}
leanbuild::set_knitr_image_path()
```

# Specific Platforms

Now we will describe some platforms that offer data storage, sharing, and analysis resources for more specific data types.

### Imaging Data Analysis Platforms

We will start with a few imaging data platforms that make it easier for researchers to perform analyses and share their data.

#### XNAT

[XNAT](https://www.xnat.org/about/) offers computing resources and tools for performing imaging analysis and for storing and sharing imaging data in a HIPAA complaint manner (more on that in the coming). Developed by the [Bukner lab](https://cnl.rc.fas.harvard.edu/) previously at the Washington University and now at Harvard, it supports a variety of imaging data as well as other data types like clinical data.  Some tools can be used with a graphical interface and others with the command-line. See [here](https://wiki.xnat.org/documentation/case-studies) for example use cases. There is also a great deal of documentation available about how to use the tools and resources available at https://wiki.xnat.org/documentation.

```{r, fig.align="center", fig.alt = "video", echo=FALSE, out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/ENk589mOkhI")
```


#### OHIF

The open health imaging foundation (OHIF) is a web-based imaging analysis platform that is widely used, particularly for radiology analysis, but it also supports whole-slide microscopy image analysis. It was developed by [Gordon Harris](https://ohif.org/team/) et al. and can be used for a variety of applications from cardiology to veterinary medicine. Check out these [example use cases](https://ohif.org/examples) of OHIF. OHIF also provides thorough documentation with images and videos about how to use the image viewer and tools available.


```{r, fig.align='center', echo = FALSE, fig.alt= "The OHIF image viewer (center) includes a tool bar for to modify the viewing and annotation of images (above center), a menu on the left to access images related to the current patient in other studies, and a menu on the right for making measurements and exporting measurement reports.", out.width= "100%"}
leanbuild::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd8d3f477a_159_2
")
```

Price: free

For those interested, Gordon Harris and others are also working on a project called [Cornerstone](https://docs.cornerstonejs.org/), with the goal of providing software for others to display medical images in web browsers.

### QuIP

### cBIBOP

### Genomic Data Analysis Platforms


### 

### Clinical Data Analysis Platforms

#### MetaboAnalyst?
https://www.metaboanalyst.ca/home.xhtml

#### mypone?
https://www.mypnoe.com/metabolic-analysis/

## WUFlux

for microbiome research https://bmcbioinformatics.biomedcentral.com/articles/10.1186/s12859-016-1314-0

Avocado - want to add in content about price, security, privacy/HIPPA, scalability/file/ram limits for each if possible.

### Machine Learning... 
want to add that tool for sharing data


https://www.toptal.com/scientific-computing/scientific-computing-with-open-source-tools

google: "https://www.youtube.com/embed/N4eT9Lfvuro"

most of these work on Virtual machines... still might need to describe this
