

```{r, include = FALSE}
ottr::set_knitr_image_path()
```

# Specific Platforms

Now we will describe some platforms that offer data storage, sharing, and analysis resources for more specific data types.

### Imaging Data Analysis Platforms

We will start with a few imaging data platforms that make it easier for researchers to perform analyses and share their data.

#### XNAT

[XNAT](https://www.xnat.org/about/) offers computing resources and tools for performing imaging analysis and for storing and sharing imaging data in a HIPAA complaint manner (more on that in the coming). Developed by the [Bukner lab](https://cnl.rc.fas.harvard.edu/) previously at the Washington University and now at Harvard, it supports a variety of imaging data as well as other data types like clinical data.  Some tools can be used with a graphical interface and others with the command-line. See [here](https://wiki.xnat.org/documentation/case-studies) for example use cases. There is also a great deal of documentation available about how to use the tools and resources available at https://wiki.xnat.org/documentation.

```{r, fig.align="center", fig.alt = "video", echo=FALSE, out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/ENk589mOkhI")
```


#### OHIF

The open health imaging foundation (OHIF) is a web-based imaging analysis platform that is widely used, particularly for radiology analysis, but it also supports whole-slide microscopy image analysis. It was developed by [Gordon Harris](https://ohif.org/team/) et al. and can be used for a variety of applications from cardiology to veterinary medicine. Check out these [example use cases](https://ohif.org/examples) of OHIF. OHIF also provides thorough documentation with images and videos about how to use the image viewer and tools available.


```{r, fig.align='center', echo = FALSE, fig.alt= "The OHIF image viewer (center) includes a tool bar for to modify the viewing and annotation of images (above center), a menu on the left to access images related to the current patient in other studies, and a menu on the right for making measurements and exporting measurement reports.", out.width= "100%"}
ottr::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gfd8d3f477a_159_2
")
```

Price: free

For those interested, Gordon Harris and others are also working on a project called [Cornerstone](https://docs.cornerstonejs.org/), with the goal of providing software for others to display medical images in web browsers.

### QuIP

```{r, fig.align="center", fig.alt = "video", echo=FALSE, out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/dK4c6ti1Dvc")
```


### IB server?


### PRISM


  The Platform for Imaging in Precision Medicine called PRISM allows users to work with the vast data available from r the Cancer Imaging Archive (TCIA).

According to Fred Prior:

>It is designed to collect, curate and manage Radiology and Pathology images, clinical information associated with those images, annotations and image derived feature sets.  PRISM is designed to run on a Kubernettes cluster, but all of the components are containerized so they can run stand-alone or in an alternate orchestration framework.


See this [article](https://ascopubs.org/doi/full/10.1200/CCI.20.00001) for more information.
 

### Genomic Data Analysis Platforms


### 

### Clinical Data Analysis Platforms

#### MetaboAnalyst?
https://www.metaboanalyst.ca/home.xhtml

#### mypone?
https://www.mypnoe.com/metabolic-analysis/

## WUFlux

for microbiome research https://bmcbioinformatics.biomedcentral.com/articles/10.1186/s12859-016-1314-0

Avocado - want to add in content about price, security, privacy/HIPAA, scalability/file/ram limits for each if possible.

### Machine Learning... 
want to add that tool for sharing data


https://www.toptal.com/scientific-computing/scientific-computing-with-open-source-tools

google: "https://www.youtube.com/embed/N4eT9Lfvuro"

most of these work on Virtual machines... still might need to describe this


## i2B2

[Informatics for Integrating Biology and the Bedside (i2b2)](https://community.i2b2.org/wiki/) is a research platform with a [web client tool](https://community.i2b2.org/wiki/display/webclient/) (which can be accessed from a browser after [downloading](https://www.i2b2.org/software/index.html) software) and data resource for healthcare related data, that has a community called the [i2b2 Community](https://community.i2b2.org/wiki/pages/viewpage.action?pageId=342684) for support. It is especially focused on pharmacology and translational research using [tranSMARt](https://pubmed.ncbi.nlm.nih.gov/24303286/) and [OpenBEL](https://www.openhealthnews.com/resources/open-biological-expression-language-openbel) translational research platforms. 

Data previously available directly with [i2b2](https://www.uab.edu/ccts/images/i2b2_FAQ.pdf) is now available as the [(n2c2) NLP research datasets](https://portal.dbmi.hms.harvard.edu/projects/n2c2-nlp/), which are data related to unstructured clinical notes at the [Partners HealthCare System](https://www.massgeneralbrigham.org/) now called Mass General Brigham from 2004 to 2014, as part of the [Department of Biomedical Informatics (DBMI) Data Portal](https://portal.dbmi.hms.harvard.edu/) in the Blavatnik Institute at Harvard Medical School. Data like this includes demographic data, visit details, immunizations etc. 

At this [link](https://community.i2b2.org/wiki/display/webclient/1.+Quick+Start) you can see a quick video introduction of what you can do with i2b2.

```{r, fig.align="center", fig.alt = "video", echo=FALSE, out.width="100%"}
knitr::include_url(" https://www.uab.edu/ccts/images/i2b2_FAQ.pdf")
```

avocado I think this should be moved to the course about clinical research.
