

```{r, include = FALSE}
leanbuild::set_knitr_image_path()
```

# Data Sharing

In this chapter we will discuss the importance of data sharing, best practices for data sharing, places where you can store your data, as well as methods to share data in contexts in which you might have thought sharing was not possible!

```{r, fig.align='center', echo = FALSE, fig.alt= "Learning Objectives: 1.Recognize general informatics computing and data management platforms. 2.Be aware of specific genomics data management systems 3.Be aware of specific imaging data management systems  4.Be aware of specific clinical data management systems", out.width="100%"}
leanbuild::include_slide("https://docs.google.com/presentation/d/1B4LwuvgA6aUopOHEAbES1Agjy7Ex2IpVAoUIoBFbsq0/edit#slide=id.gf5f8818810_1_171")
```

## Data sharing is important!

Sharing data is critical for optimizing the advancement of scientific understanding. Now that labs all over the world are producing massive amounts of data, there are many discoveries that can be made by just using this existing data. 

This is so important, that starting in January, 2023 the NIH will require specific sharing practices for data management and sharing. See the announcement [here](https://grants.nih.gov/grants/guide/notice-files/NOT-OD-21-013.html).


AVOCADO Somewhere want to talk about where you can access data... or share it...
https://transcriptomics.sequentiabiotech.com/
Repositories: GEO Gene Expression Omnibus(GEO), SRA(Sequencing Read Archive), ArrayExpress
